---
title: "Bayesian Intensity Modeling"
format: html
editor: visual
---

```{r}
#| message: false
#| warning: false


# load libraries
library(ggplot2)
library(ggridges)
library(ggforce)
library(ggpubr)

library(tidyverse)
library(purrr)

library(tidybayes)
library(brms)

library(here)

library(rcarbon)
library(c14bazAAR)

library(sf)
library(terra)
library(rnaturalearth)
library(rnaturalearthdata)

library(cmdstanr)
library(posterior)
library(bayesplot)


library(dbscan)
library(factoextra)

library(akima)

require(doSNOW)
require(doParallel)
require(doRNG)
```

```{r}
seed = 1
set.seed(1)
# Load data from southern African database

# set directory
dir = here::here()

# read in and modify data names and assign as date list for c14bazAAR
c14_za <- read_csv(paste0(dir, "/data/SARD_Mar2021_14C.csv")) |>
  rename(c14age = Date, c14std = Uncertainty,  labnr = 'Lab ID',
         lon = DecdegE, lat = DecdegS,
         material = 'Material dated', species = Species) |>
  as.c14_date_list()

# load the SASSI database for full ZA site locations
sassi <- read_csv(paste0(dir, '/data/SASSI_v3.2_S1.csv')) |>
  pivot_longer('ESA':'CeramicLSA', names_to = 'Industry', 
               values_to = 'Presence') |>
  mutate(Presence = ifelse(!is.na(Presence), 1, 0)) |>
  filter((Industry %in% c('LSA', 'GenLSA', 'Robberg', 'Oakhurst', 
                         'Wilton', 'FinalLSA', 'CeramicLSA')) &
           (Presence == 1))


```

```{r}
# filter by late pleistocene
max_age <- 21000 
min_age <- 2000
temp_res <- 100

#filter to remove human remains as they produce errors when calibrating
za <- c14_za |>
  filter(species != "Homo sapiens" | is.na(species)) |>
  
  # filter by dates older than 200 years and less than 50 ka
  filter(c14age > min_age, c14age < max_age)


# filter by terrestrial material and apply the terrestrial calibration curve
za_terr <- za |>
  filter(!material %in% c('Marine shell', 'marine shell',
                          'marine other', 'iron')) |>
  drop_na(c14age, c14std)

za_terr_cal <- rcarbon::calibrate(za_terr$c14age , za_terr$c14std,
                                      calCurves = 'shcal20',
                                      ids = za_terr$labnr)

terr_bins <- binPrep(za_terr$Site, ages = za_terr_cal, h = 250)

# filter by marine material and apply marien curve and offset
za_marine <- za |>
  filter(material %in% c('Marine shell', 'marine shell',
                         'marine other')) |>
  drop_na(c14age, c14std)

za_marine_cal <- rcarbon::calibrate(za_marine$c14age, 
                                    za_marine$c14std,
                                    calCurves = rep('marine20',
                                                    length(za_marine$labnr)),
                                    resOffset = rep(187, length(za_marine$labnr)),
                                    resErrors = rep(18, length(za_marine$labnr)),
                                    ids = za_marine$labnr)

marine_bins <- binPrep(za_marine$Site, ages = za_marine_cal, h = 250)

# combine data
za_cal <- rcarbon::combine(za_terr_cal, za_marine_cal)
```

```{r}
time_range =  c(max_age, min_age)
time_range_df = data.frame(calBP = max_age:min_age)



za_cal_site <- za_cal |>
  as_tibble()

# Need to sample sites and dates, then combine
za_spd_df <- tibble(site = character(),
                    calBP = numeric(),
                    PrDens = numeric())

for(i in unique(za$labnr)){
  

    
    
    tmp_site <- za |>
      filter(labnr == i) |>
      select(Site)
    
    tmp_za <- za |>
      filter(Site == tmp_site$Site)
    
    
    za_terr <- tmp_za |>
      filter(!material %in% c('Marine shell', 'marine shell',
                              'marine other', 'iron')) |>
      drop_na(c14age, c14std)
    
    if(dim(za_terr)[1] != 0){
    
    za_terr_cal <- rcarbon::calibrate(za_terr$c14age , za_terr$c14std,
                                      calCurves = 'shcal20',
                                      ids = za_terr$labnr,
                                      verbose = FALSE)
    
    }
    
    # filter by marine material and apply marien curve and offset
    za_marine <- tmp_za |>
      filter(material %in% c('Marine shell', 'marine shell',
                             'marine other')) |>
      drop_na(c14age, c14std)
    if(dim(za_marine)[1] != 0){
    
      za_marine_cal <- rcarbon::calibrate(za_marine$c14age, 
                                          za_marine$c14std,
                                          calCurves = rep('marine20',
                                                          length(za_marine$labnr)),
                                          resOffset = rep(187, length(za_marine$labnr)),
                                          resErrors = rep(18, length(za_marine$labnr)),
                                          ids = za_marine$labnr,
                                          
                                          verbose = FALSE)
    }
    
    # combine data
    if((dim(za_terr)[1] == 0 & dim(za_marine)[1] != 0)){
      za_cal <- za_marine_cal
    } else if((dim(za_terr)[1] != 0 & dim(za_marine)[1] == 0)){
      za_cal <- za_terr_cal
    } else{
      za_cal <- rcarbon::combine(za_terr_cal, za_marine_cal)
    }
    
    
    target_bins = binPrep(sites = c(za_terr$Site, za_marine$Site),
                          ages = c(za_terr$c14age, za_marine$c14age), h = 100)
    
    
    tmp_spd = spd(x = za_cal,
                  bins = target_bins, 
                  timeRange = c(max_age, min_age), 
                  datenormalised = FALSE,
                  runm = 100)
    
    tmp_spd_df <- as_tibble(tmp_spd$grid)
    
    za_spd_df <- za_spd_df |>
      add_row(site = tmp_site$Site,
              calBP = tmp_spd_df$calBP,
              PrDens = tmp_spd_df$PrDens)
}
```

```{r}
site_prob <- za_spd_df |>
  mutate(
    bin = floor(calBP / temp_res) * temp_res
  ) |>
  group_by(site, bin) |>
  summarise(
    prob = sum(PrDens),
    .groups = "drop"
  ) |>
  complete(
    site = unique(za_spd_df$site),
    bin = seq(
      floor(max_age / temp_res) * temp_res,
      floor(min_age / temp_res) * temp_res,
      by = -temp_res
    ),
    fill = list(prob = 0)
  ) |>
  group_by(site) |>
  mutate(
    p = prob / sum(prob)
  ) |>
  ungroup()



dates_samp <- site_prob |>
  group_by(site) |>
  filter(sum(p) > 0) |>
  slice_sample(
    n = 1,
    weight_by = p,
    replace = TRUE
  ) |>
  ungroup() |>
  rename(calBP = bin)


iter <- 5000
seed = seed + 1
set.seed(seed)

dates_samp <- purrr::map_dfr(
  seq_len(iter),
  function(k) {
    
    site_prob |>
      group_by(site) |>
      filter(sum(p) > 0) |>
      slice_sample(
        n = 1,
        weight_by = p,
        replace = TRUE
      ) |>
      ungroup() |>
      mutate(iter = k)
    
  }
)



site_counts <- site_prob |>
  select(site, bin, p) |>
  left_join(
    dates_samp |>
      count(site, bin, name = "site_count"),
    by = c("site", "bin")
  ) |>
  mutate(
    site_count = replace_na(site_count, 0)
  )


local_lambda <- site_prob |>
  mutate(
    lambda = iter * p,
    local_rate = lambda / temp_res
  )


za_pois_all <- local_lambda |>
  mutate(
    count = rbinom(
      n(),
      size = iter,
      prob = p
    )
  ) |>
  rename(calBP = bin)
```

```{r}
za_pois_all |>
  rename(Count = count) |>
  ggplot(aes(x = calBP, y = Count, fill = Count))+
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "darkorange", high = "darkblue") +
  xlab("Calibrated Years BP")+
  scale_x_reverse()+
  theme_minimal(base_size = 8)+
  theme(axis.text.x = element_text(angle = 45,  hjust = 1))
```

```{r}
###########################################
########### ASSIGN SARD DATA ##############
za_comb <- za_pois_all |>
  rename(Site = site) |>
  left_join(za |> select(Site, lat, lon, Country, Biome, `Archaeological Period`,
                          `Archaeological Sub-chronology`, `Province or district`,
                          `Site Type`) |> distinct(Site, .keep_all = TRUE),
             by = 'Site') |>
  ungroup()


sites_sf <- st_as_sf(
  za_comb,
  coords = c("lon", "lat"),
  crs = 4326
)

sites_proj <- st_transform(
  sites_sf,
  32734
)


coords <- st_coordinates(sites_proj)

za_coords <- za_comb |>
  mutate(x = coords[,1],
         y = coords[,2])

study_area <- st_bbox(sites_proj) |>
  st_as_sfc() |>
  st_buffer(dist = 10000) |>
  vect()

###########################################
########### ASSIGN SASSI DATA ##############


sassi_sf <- st_as_sf(
  sassi,
  coords = c("Longitude", "Latitude"),
  crs = 4326
)

sassi_proj <- st_transform(
  sassi_sf,
  32734
)


sassi_coords <- st_coordinates(sassi_proj)

sassi_coords <- sassi |>
  mutate(x = sassi_coords[,1],
         y = sassi_coords[,2])

sassi_area <- st_bbox(sassi_proj) |>
  st_as_sfc() |>
  st_buffer(dist = 10000) |>
  vect()

sassi_holocene <- sassi_proj |>
  mutate(x = st_coordinates(sassi_proj)[,1],
         y = st_coordinates(sassi_proj)[,2],
         
         Industry = as.factor(ifelse(Industry == 'CeramicLSA', 
                           'Ceramic Final Later Stone Age',
                    ifelse(Industry == 'FinalLSA', 'Final Later Stone Age',
                           Industry))),
         
         Industry = fct_relevel(Industry, c('Robberg', 'Oakhurst',
                                            'Wilton', 'Final Later Stone Age',
                                            'Ceramic Final Later Stone Age')))

africa_map <- ne_countries(scale = "medium", 
                           continent = "Africa", returnclass = "sf")


png("map1.png", width = 1200, height = 800, res = 300)

ggplot()+
  geom_sf(data = africa_map, fill = "antiquewhite", color = "darkgray") +
  geom_sf(data = sassi_holocene  |>
            filter(Industry != 'GenLSA', Industry != 'LSA'),
          aes(color = Biome), size = 0.45) +
  facet_wrap(~Industry)+
  coord_sf(xlim = c(10, 42), ylim = c(-35, -10), expand = FALSE) +
  theme_minimal() +
  theme(axis.text = element_text(size = 3),
        axis.title = element_text(size = 3),
        strip.text = element_text(size = 4),
        
        legend.title = element_text(size = 5),
        legend.text = element_text(size = 4))

dev.off()
```

```{r}
za_holocene <- sites_proj |>
  mutate(x = st_coordinates(sites_proj)[,1],
         y = st_coordinates(sites_proj)[,2])

africa_map <- ne_countries(scale = "medium", 
                           continent = "Africa", returnclass = "sf")


ggplot()+
  geom_sf(data = africa_map, fill = "antiquewhite", color = "darkgray") +
  geom_sf(data = za_holocene, aes(color = `Biome`)) +
  coord_sf(xlim = c(10, 42), ylim = c(-35, -10), expand = FALSE) +
  theme_minimal()
```

```{r}
# 1. Prepare your data (using built-in multishapes dataset example)

df_tmp <- sassi_holocene[, 21:22] |>
  st_drop_geometry() |>
  distinct()


kNNdistplot(df_tmp, k = 4)
abline(h = 105000, col = "red", lty = 2)
```

```{r}
nn <- frNN(sassi_holocene |> distinct(Site, .keep_all = TRUE) |> st_coordinates(),
           eps = 105000)

hist(lengths(adjacencylist(nn)),
     xlab = "k", main="Number of Neighbors",
     sub = paste("Neighborhood size eps =", nn$eps))


min_obs <- 4


clustered_data <- sassi_holocene |>
  distinct(Site, .keep_all = TRUE) |>
  select(Site) |>
  
  mutate(cluster = {
    
    # Run DBSCAN (adjust eps and minPts to your data's scale)
    res <- dbscan(nn, minPts = min_obs)
    
    # Return cluster assignments (0 represents noise/outliers)
    res$cluster
  }) |>
  st_drop_geometry() |>
  
  full_join(sassi_holocene, by = "Site") |>
  st_sf()

clustered_data |>
  group_by(cluster) |>
  mutate(n_cluster = n()) |>
  filter(n_cluster < 15) |>
  group_by(Site, cluster, n_cluster) |>
  summarize(n_site = n()) |>
  mutate(Site_ex = ifelse((n_cluster == n_site) | (n_site < min_obs), 1, 0)) |>
  filter(Site_ex == 1) -> site_exc

#  Remove the noise cluster and clusters with too few site-specific observations
clustered_data <- clustered_data |> filter(!(cluster %in% c(0, site_exc$cluster)))


hulls_sf <- clustered_data |>
  group_by(cluster) |>
  summarize(geometry = st_combine(geometry)) |>
  st_convex_hull()


png("map2.png", width = 1200, height = 800, res = 300)
# Explore clustered data
ggplot()+
  geom_sf(data = africa_map, fill = "antiquewhite", color = "darkgray") +
  #geom_sf(data = hulls_sf, aes(fill = factor(cluster)), 
   #       alpha = 0.2, color = "black", size = 0.5) +
  geom_sf(data = clustered_data, aes(color = factor(cluster)), size = 0.45) +
  geom_mark_hull(alpha = 0.2, expand = unit(3, "mm")) +
  scale_color_discrete(name = "Cluster",
                       labels = c(
                         '1' = '1 - Southern Coast',
                         
                         '2' = '2 - Northwest Coast',
                         '9' = '9 - Northwest Coast',
                         
                         '3' = '3 - Interior Grassland',
                         
                         '4' = '4 - East Coast',
                         
                         '5' = '5 - Interior Karoo',
                         
                         '6' = '6 - Northeast',
                         '7' = '7 - Northeast',
                         '8' = '8 - Northeast'
                       )) +
  coord_sf(xlim = c(10, 42), ylim = c(-35, -10), expand = FALSE) +
  theme_minimal()+
  theme(axis.text = element_text(size = 3),
        axis.title = element_text(size = 3),
        strip.text = element_text(size = 4),
        
        legend.title = element_text(size = 5),
        legend.text = element_text(size = 4))

dev.off()
```

# Hierarchical Modeling

```{r}

fit_df <- za_holocene |>
  st_drop_geometry() |>
  rename(site_count = count,
         Tech = `Archaeological Sub-chronology`) |>
  ungroup() |>
  
  mutate(time_s = as.numeric(scale(calBP)),
         time_red = calBP/1000,
         time_ka = -(calBP - mean(calBP, na.rm = TRUE))/1000,
         
         Tech = ifelse(str_detect(Tech, 'Iron'), 'Ceramic LSA', Tech),
         
         iter = iter) |>
  filter(Tech != 'ELSA', Tech != 'Final MSA')




fit_df |>
  ggplot(aes(x = calBP, y = site_count, fill = site_count))+
  geom_bar(stat = "identity") +
  facet_wrap(~Tech, scale = "free") +
  scale_fill_gradient(low = "darkorange", high = "darkblue") +
  xlab("Calibrated Years BP")+
  theme_minimal(base_size = 8)+
  theme(axis.text.x = element_text(angle = 45,  hjust = 1))
```

## Bayesian Modeling

```{r}

# this model shows cluster show almost no variation after accounting for mean site count
#seed <- seed + 1
seed <- 8
fit_time <- brm(
  site_count ~
    1 + time_ka,
  
  family = negbinomial(),
  
  data = fit_df,
  
  prior = c(
    prior(normal(0, 2), class = "Intercept")
  ),
  
  chains = 1,
  iter = 4000,
  warmup = 2000,
  
  control = list(
    adapt_delta = 0.95
  ),
  
  cores = 4,
  threads = threading(8),
  backend = 'cmdstanr',
  seed = seed
)

seed <- 9
fit_time2 <- brm(
  site_count ~
    1 + time_ka + (1  + time_ka | Tech),
  
  family = negbinomial(),
  
  data = fit_df,
  
  prior = c(
    prior(normal(0, 2), class = "Intercept")
  ),
  
  chains = 1,
  iter = 4000,
  warmup = 2000,
  
  control = list(
    adapt_delta = 0.95
  ),
  
  cores = 4,
  threads = threading(8),
  backend = 'cmdstanr',
  seed = seed
)


```

# Results

## Hypotheses

### $H_1$: ***Rates of radiocarbon intensity vary between archaeological sites in southern Africa.***

### $H_2$: ***Periods that correspond to the technological industries will show different rates of radiocarbon intensity based on varying population sizes.***

#### $H_{2a}$: The period associated with Robberg technology will show reduced radiocarbon intensity compared with the Oakhurst, Wilton, and Post-Wilton technological periods.

#### $H_{2b}$: The period associated with Oakhurst technology should show higher radiocarbon intensity rates than the Robberg technological period.

#### $H_{2c}$: The period associated with Wilton technology should show lower radiocarbon intensity rates than the Oakhurst technological period.

#### $H_{2D}$: The period associated with Post-Wilton technology should show higher radiocarbon intensity rates than the Wilton technological period.

### $H_3$: ***Radiocarbon intensity for technological periods associated with greater population sizes will have regional-distinct patterns but will be regionally similar during periods with low population sizes.***

#### $H_{3a}$: During the Robberg period, regional radiocarbon intensity will be similar across southern Africa.

#### $H_{3B}$: During the Oakhurst period, regional radiocarbon intensity will be different across southern Africa.

#### $H_{3C}$: During the Wilton period, regional radiocarbon intensity will be similar across southern Africa.

#### $H_{3D}$: During the Post-Wilton period, regional radiocarbon intensity will be different across southern Africa.

## Results

### 1) Do rates of radiocarbon intensity vary through time in southern Africa?

Yes, the radiocarbon intensity increases exponentially towards the present. The mean rate is 0.96 for Robberg period, 9.63 for the Oakhurst period, 31.4 for the WIlton period, and 78.5 for the Final LSA period.

```{r}
# compute posterior slope estimates for the first model including time
time_mean <- mean(fit_df$calBP, na.rm = TRUE)

prediction_grid <- expand_grid(
  calBP = seq(min_age, max_age, by = 1000)
) |>
  mutate(time_ka = -(calBP - time_mean)/1000)



draws <- posterior_epred(
  fit_time,
  newdata = prediction_grid
  ) |>
  as_tibble() |>
  rename_with(
    .fn = ~ paste0(unique(prediction_grid$calBP)),
    .cols = starts_with('V')
  ) |>
  pivot_longer(
    cols = '2000':'21000',
    values_to = 'estimate',
    names_to = 'time'
  ) |>
  mutate(time = as.numeric(time)) |>
  group_by(time) |>
  
  summarise(
    slope = median(estimate),
    lower = quantile(estimate, 0.05),
    upper = quantile(estimate, 0.95),
    
    .groups = "drop"
  )
```

```{r}
draws |>
  ggplot(aes(x = time, y = slope)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.75) +
  scale_x_reverse() +
  
  annotate("rect", xmin = 19923, xmax = 13145, ymin = -Inf, ymax = Inf,
           fill = "blue", alpha = 0.2) +
  annotate("text", x = (19923+13145)/2, y = 25, label = "Robberg", 
           color = "blue", fontface = "bold", vjust = -0.5) +
  
  annotate("rect", xmin = 12789, xmax = 8705, ymin = -Inf, ymax = Inf,
           fill = "forestgreen", alpha = 0.2) +
  annotate("text", x = (12789+8705)/2, y = 25, label = "Oakhurst", 
            color = "forestgreen", fontface = "bold", vjust = -0.5) +
  
  annotate("rect", xmin = 7784, xmax = 4636, ymin = -Inf, ymax = Inf,
           fill = "orange", alpha = 0.2) +
  annotate("text", x = (7784+4636)/2, y = 25, label = "Wilton", 
            color = "orange", fontface = "bold", vjust = -0.5) +
  
  annotate("rect", xmin = 3898, xmax = 1318, ymin = -Inf, ymax = Inf,
           fill = "gold", alpha = 0.2) +
  annotate("text", x = (3898+1318)/2, y = 19, label = "Final Later\nStone Age", 
            color = "gold", fontface = "bold", vjust = -0.5) +
  
  
  ylab("Rate of Radiocarbon Intesnity/1,000 years") +
  xlab("Time (yrs cal BP)") +
  theme_minimal()
```

### 2) Do rates of radiocarbon intensity vary through time and technological classification in southern Africa?

No, a pairwise comparison shows that the assigned technological classes do not change through time within southern Africa. This is likely because the actual classifications begin prior to and after the proposed temporal interval Lombard et al. (2022) assigned.

```{r}
# compute posterior slope estimates for each cluster
time_mean <- mean(fit_df$calBP, na.rm = TRUE)

prediction_grid <- expand_grid(
  calBP = seq(min_age, max_age, by = 1000),
  Tech = unique(fit_df$Tech)
) |>
  mutate(time_ka = -(calBP - time_mean)/1000) |>
  filter(!is.na(Tech))

draws <- posterior_epred(
  fit_time2,
  newdata = prediction_grid
)  |>
  as_tibble() |>
  rename_with(
    .fn = ~ unique(paste0(prediction_grid$calBP, "_", 
                          prediction_grid$Tech)),
    .cols = starts_with('V')
  ) |>
  pivot_longer(
    cols = everything(),
    values_to = 'estimate',
    names_to = 'time'
  ) |>
  separate_wider_delim(
    cols = time, 
    delim = "_", 
    names = c("time", "tech")
  ) |>
  mutate(time = as.numeric(time)) |>
  group_by(time, tech) |>
  
  summarise(
    slope = median(estimate),
    lower = quantile(estimate, 0.05),
    upper = quantile(estimate, 0.95),
    
    .groups = "drop"
  ) 


draws |>
  filter(tech != 'Ceramic LSA') |>
  mutate(tech = factor(tech, 
                          levels = c('Robberg', 'Oakhurst',
                                     'Wilton', 'Final LSA'))) |>
  
  ggplot(aes(x = time, y = slope)) +
  geom_smooth(method = "loess", se = TRUE, span = 0.75) +
  scale_x_reverse() +
  
  annotate("rect", xmin = 19923, xmax = 13145, ymin = -Inf, ymax = Inf,
           fill = "blue", alpha = 0.2) +
  annotate("text", x = (19923+13145)/2, y = 25, label = "Robberg", 
           color = "blue", fontface = "bold", vjust = -0.5) +
  
  annotate("rect", xmin = 12789, xmax = 8705, ymin = -Inf, ymax = Inf,
           fill = "forestgreen", alpha = 0.2) +
  annotate("text", x = (12789+8705)/2, y = 25, label = "Oakhurst", 
           color = "forestgreen", fontface = "bold", vjust = -0.5) +
  
  annotate("rect", xmin = 7784, xmax = 4636, ymin = -Inf, ymax = Inf,
           fill = "orange", alpha = 0.2) +
  annotate("text", x = (7784+4636)/2, y = 25, label = "Wilton", 
           color = "orange", fontface = "bold", vjust = -0.5) +
  
  annotate("rect", xmin = 3898, xmax = 1318, ymin = -Inf, ymax = Inf,
           fill = "gold", alpha = 0.2) +
  annotate("text", x = (3898+1318)/2, y = 19, label = "Final Later\nStone Age", 
           color = "gold", fontface = "bold", vjust = -0.5) +
  
  facet_wrap(~tech, scales = 'free') +
  
  
  ylab("Rate of Radiocarbon Intesnity/1,000 years") +
  xlab("Time (yrs cal BP)") +
  theme_minimal()


```

```{r}
# examine the radiocarbon rate by technology
draws_tech <- posterior_epred(
  fit_time2,
  newdata = prediction_grid
)  |>
  as_tibble() |>
  rename_with(
    .fn = ~ unique(paste0(prediction_grid$calBP, "_", 
                          prediction_grid$Tech)),
    .cols = starts_with('V')
  ) |>
  pivot_longer(
    cols = everything(),
    values_to = 'estimate',
    names_to = 'time'
  ) |>
  separate_wider_delim(
    cols = time, 
    delim = "_", 
    names = c("time", "tech")
  ) |>
  mutate(time = as.numeric(time),
         tech = as.factor(tech)) |>
  group_by(tech) |>
  #filter(time <7784) |> # filter this for the wilton and final LSA comparison
  
  summarise(
    slope = median(estimate),
    lower = quantile(estimate, 0.05),
    upper = quantile(estimate, 0.95),
    
    .groups = "drop"
  ) 


# examine the radiocarbon rate by period
draws_period <- posterior_epred(
  fit_time2,
  newdata = prediction_grid
)  |>
  as_tibble() |>
  rename_with(
    .fn = ~ unique(paste0(prediction_grid$calBP, "_", 
                          prediction_grid$Tech)),
    .cols = starts_with('V')
  ) |>
  pivot_longer(
    cols = everything(),
    values_to = 'estimate',
    names_to = 'time'
  ) |>
  separate_wider_delim(
    cols = time, 
    delim = "_", 
    names = c("time", "tech")
  ) |>
  mutate(time = as.numeric(time),
         
         period = ifelse(time > 13000, 'Robberg',
                         ifelse(time > 8000, 'Oakhurst',
                                ifelse(time > 4000, 'Wilton',
                                       ifelse(time > 1300, 'Final LSA',
                                              'Ceramic LSA')))),
         tech = factor(tech)) |>
  group_by(period) |>
  
  summarise(
    slope = median(estimate),
    lower = quantile(estimate, 0.05),
    upper = quantile(estimate, 0.95),
    
    .groups = "drop"
  ) 

# examine the radiocarbon rate bewtween time period and tehnology

draws_tech_period <- posterior_epred(
  fit_time2,
  newdata = prediction_grid
)  |>
  as_tibble() |>
  rename_with(
    .fn = ~ unique(paste0(prediction_grid$calBP, "_", 
                          prediction_grid$Tech)),
    .cols = starts_with('V')
  ) |>
  pivot_longer(
    cols = everything(),
    values_to = 'estimate',
    names_to = 'time'
  ) |>
  separate_wider_delim(
    cols = time, 
    delim = "_", 
    names = c("time", "tech")
  ) |>
  mutate(time = as.numeric(time),
         
         period = ifelse(time > 13000, 'Robberg',
                         ifelse(time > 8000, 'Oakhurst',
                                ifelse(time > 4000, 'Wilton',
                                       ifelse(time > 1300, 'Final LSA',
                                              'Ceramic LSA')))),
         tech = factor(tech)) |>
  group_by(tech, period) |>
  
  summarise(
    slope = median(estimate),
    lower = quantile(estimate, 0.05),
    upper = quantile(estimate, 0.95),
    
    .groups = "drop"
  ) 



```

```{r}
ep <- posterior_epred(
  fit_time2,
  newdata = prediction_grid
) |> 
  as_tibble() |>
  mutate(draw = row_number()) |>
  pivot_longer(
    cols = -draw,
    names_to = "pred_id",
    values_to = "estimate"
  ) |>
  mutate(
    pred_id = as.integer(sub("V", "", pred_id)),
    time = prediction_grid$calBP[pred_id],
    tech = prediction_grid$Tech[pred_id]
  ) |>
  mutate(
    time = as.numeric(time),
    tech = factor(tech)
  ) |>
  pivot_wider(
    id_cols = c(draw, time),
    names_from = tech,
    values_from = estimate
  )

# Evaluate the probability that the radiocarbon intensity rates are different between technological classifications
ep |>
  summarize(`Robberg > Oakhurst` = mean(Robberg > Oakhurst),
            `Oakhurst > Wilton` = mean(Oakhurst > Wilton),
            `Wilton > Final LSA` = mean(Wilton > `Final LSA`),
            
            `Robberg > Wilton` = mean(Robberg > `Wilton`))
```

### 3) Do different technological periods correspond to different radiocarbon intensities across southern African regions?

Yes, of particular interest is the fact that the interior-karoo radiocarbon intensity increases into the later Holocene, suggesting a continuous occupation of this region during the middle and later Holocene despoit previous predictions. However, the pattern is much different when we consider the radiocarbon intensity associated to sites assigned to distinct technological classifications.

In particular, though the interior-karoo was significantly different between the Robberg and Oakhurst period (19 ka cal BP - 8.7 ka cal BP), there is no significant difference bewteen radiocarbon intensity and the sites that show Robberg and Oakhurst technologies. However, Wilton technology has a higher rate of radiocarbon intensity within the interior-karoo, implying greater population around sites associated with wilton technology in the interior-karoo. The final LSA also has a lower rate in the interior-karoo than does wilton technology. Given the graphic, this is the general trend for all of southern Africa.

**In conclusion, the technological patterns show homogeneous patterns through southern Africa whereas the temporal trends show significant differences across southern Africa.** Each period shows a monotonic increase in radiocarbon intensity through time, but Africa is regionally distinct over specific periods of time, which the technological trends do not track.

```{r}
# combine clustered data from SASSI to the SARD database
df_tmp <- za_holocene |>
  full_join(clustered_data |> 
              st_drop_geometry() |> 
              select(Site, cluster) |>
              distinct(), by = "Site") |>
  filter(!is.na(cluster))

fit_df <- df_tmp |>
  st_drop_geometry() |>
  rename(site_count = count,
         Tech = `Archaeological Sub-chronology`) |>
  group_by(cluster, calBP) |>
  mutate(cluster_dates = n()) |>
  ungroup() |>
  
  mutate(time_s = as.numeric(scale(calBP)),
         time_red = calBP/1000,
         time_ka = -(calBP - mean(calBP, na.rm = TRUE))/1000,
         
         cluster_fct = factor(cluster),
         cluster_num = as.numeric(cluster),
         cluster_cat = ifelse(cluster %in% c(9, 2), 'Northwest Coast',
                              ifelse(cluster %in% c(8, 7, 6), 'Northeast',
                                     ifelse(cluster %in% c(5), 'Interior-Karoo',
                                            ifelse(cluster %in% c(3), 'Interior-Grassland',
                                                   ifelse(cluster == 4, 'East Coast', 'Southern Coast'))))),
         
         Tech = ifelse(str_detect(Tech, 'Iron'), 'Ceramic LSA', Tech),
         
         iter = iter) |>
  filter(Tech != 'ELSA', Tech != 'Final MSA')


png("map3.png", width = 1200, height = 800, res = 300)

ggplot()+
  geom_sf(data = africa_map, fill = "antiquewhite", color = "darkgray") +
  geom_sf(data = df_tmp, aes(color = factor(cluster)), size = 0.45) +
  geom_mark_hull(alpha = 0.2, expand = unit(3, "mm")) +
  scale_color_discrete(name = "Cluster",
                       labels = c(
                         '1' = '1 - Southern Coast',
                         
                         '2' = '2 - Northwest Coast',
                         '9' = '9 - Northwest Coast',
                         
                         '3' = '3 - Interior Grassland',
                         
                         '4' = '4 - East Coast',
                         
                         '5' = '5 - Interior Karoo',
                         
                         '6' = '6 - Northeast',
                         '7' = '7 - Northeast',
                         '8' = '8 - Northeast'
                       )) +
  coord_sf(xlim = c(10, 42), ylim = c(-35, -10), expand = FALSE) +
  theme_minimal()+
  theme(axis.text = element_text(size = 3),
        axis.title = element_text(size = 3),
        strip.text = element_text(size = 4),
        
        legend.title = element_text(size = 5),
        legend.text = element_text(size = 4))

dev.off()
```

```{r}
############################################################
#seed <- seed + 1
seed <- 10
fit_space_h3 <- brm(
  site_count ~
    1 + time_ka + (0 + time_ka || cluster_cat),
  
  family = negbinomial(),
  
  data = fit_df,
  
  prior = c(
    prior(normal(0, 2), class = "Intercept")
  ),
  
  chains = 1,
  iter = 4000,
  warmup = 2000,
  
  control = list(
    adapt_delta = 0.95
  ),
  
  cores = 4,
  threads = threading(8),
  backend = 'cmdstanr',
  seed = seed
)


############################################################
#seed <- seed + 1
seed <- 11
fit_space_h4 <- brm(
  site_count ~
    1 + time_ka + (1 + time_ka | Tech) + (0 + time_ka || cluster_cat),
  
  family = negbinomial(),
  
  data = fit_df,
  
  prior = c(
    prior(normal(0, 2), class = "Intercept")
  ),
  
  chains = 1,
  iter = 4000,
  warmup = 2000,
  
  control = list(
    adapt_delta = 0.95
  ),
  
  cores = 4,
  threads = threading(8),
  backend = 'cmdstanr',
  seed = seed
)
```

```{r}
# Compare age ranges
intervals <- tribble(
  ~period, ~start, ~end,
  "Robberg", 13145, 19923,
  "Oakhurst",   8705, 12789,
  "Wilton",    4636,  7784,
  "Final LSA", 1318, 3898
) |>
  mutate(period = as.factor(period),
         period = fct_relevel(period, c('Robberg', 'Oakhurst', 'Wilton',
                                        'Final LSA')))


time_mean <- mean(fit_df$calBP, na.rm = TRUE)

interval_grid <- crossing(
  intervals,
  cluster_cat = unique(fit_df$cluster_cat),
  calBP = seq(1000, 21000, by = 500)
) |>
  filter(
    calBP >= pmin(start, end),
    calBP <= pmax(start, end)
  ) |>
  mutate(
    time_ka = -(calBP - time_mean)/1000
  )

interval_pred <- posterior_epred(
  fit_space_h3,
  newdata = interval_grid,
  re_formula = NULL
)


pred_long <- interval_grid |>
  select(period, cluster_cat, calBP) |>
  bind_cols(
    as.data.frame(t(interval_pred))
  ) |>
  pivot_longer(
    cols = starts_with("V"),
    names_to = "draw",
    values_to = "mu"
  )

# caculate expectations
interval_summary <- pred_long |>
  group_by(period, cluster_cat, draw) |>
  mutate(period = as.factor(period)) |> 
  summarise(
    mean_count = mean(mu),
    .groups = "drop"
  ) |>
  group_by(period, cluster_cat) |>
  summarise(
    estimate = median(mean_count),
    lower = quantile(mean_count, 0.05),
    upper = quantile(mean_count, 0.95),
    .groups = "drop"
  )

interval_summary |>
  
  ggplot(aes(x = cluster_cat, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.15
  ) +
  facet_wrap(~period, scales = "free_y") +
  ylab("Radiocarbon Intensity") +
  xlab("Regional Aggregate") +
  theme_minimal()+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) -> gg_1


interval_summary |>
  
  ggplot(aes(x = period, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.15
  ) +
  facet_wrap(~cluster_cat, scales = "free_y") +
  ylab("Radiocarbon Intensity") +
  xlab("Regional Aggregate") +
  theme_minimal()+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) -> gg_2

gg_1
```

```{r}

draws <- as_draws_df(fit_space_h3)

slope_draws <- draws |>
  select(
    .draw,
    b_time_ka,
    starts_with("r_cluster_cat[")
  ) |>
  pivot_longer(
    cols = starts_with("r_cluster_cat["),
    names_to = "cluster",
    values_to = "slope_deviation"
  ) |>
  mutate(
    cluster_slope = b_time_ka + slope_deviation,
    cluster = str_extract(cluster, "(?<=\\[)[^,]+")
  )  


slope_pairs <- combn(
  unique(slope_draws$cluster),
  2,
  simplify = FALSE
)

pairwise_slopes <- map_dfr(
  slope_pairs,
  \(x) {
    
    diff <- slope_draws |>
      filter(cluster %in% x) |>
      select(.draw, cluster, cluster_slope) |>
      pivot_wider(
        names_from = cluster,
        values_from = cluster_slope
      ) |>
      mutate(
        difference = .data[[x[1]]] - .data[[x[2]]]
      ) |>
      pull(difference)
    
    tibble(
      cluster_1 = x[1],
      cluster_2 = x[2],
      estimate = median(diff),
      lower = quantile(diff, 0.025),
      upper = quantile(diff, 0.975),
      prob_cluster1_greater = mean(diff > 0),
      
      # optional: ratio of temporal rate multipliers
      rate_ratio = median(exp(diff)),
      rate_ratio_lower = quantile(exp(diff), 0.025),
      rate_ratio_upper = quantile(exp(diff), 0.975)
    )
  }
)

pairwise_slopes |>
  filter(lower > 0 | upper < 0)
```

```{r}
# Compare age ranges
intervals <- tribble(
  ~period, ~start, ~end,
  "Robberg", 13145, 19923,
  "Oakhurst",   8705, 12789,
  "Wilton",    4636,  7784,
  "Final LSA", 1318, 3898
) |>
  mutate(period = as.factor(period),
         period = fct_relevel(period, c('Robberg', 'Oakhurst', 'Wilton',
                                        'Final LSA')))


time_mean <- mean(fit_df$calBP, na.rm = TRUE)

interval_grid <- crossing(
  intervals,
  cluster_cat = unique(fit_df$cluster_cat),
  Tech = unique(fit_df$Tech),
  calBP = seq(1000, 21000, by = 500)
) |>
  filter(
    calBP >= pmin(start, end),
    calBP <= pmax(start, end)
  ) |>
  mutate(
    time_ka = -(calBP - time_mean)/1000
  )

interval_pred <- posterior_epred(
  fit_space_h4,
  newdata = interval_grid,
  re_formula = NULL
)


pred_long <- interval_grid |>
  select(cluster_cat, Tech, calBP) |>
  bind_cols(
    as.data.frame(t(interval_pred))
  ) |>
  pivot_longer(
    cols = starts_with("V"),
    names_to = "draw",
    values_to = "mu"
  )

# caculate expectations
interval_summary <- pred_long |>
  group_by(Tech, cluster_cat, draw) |>
  mutate(Tech = as.factor(Tech)) |> 
  summarise(
    mean_count = mean(mu),
    .groups = "drop"
  ) |>
  group_by(Tech, cluster_cat) |>
  summarise(
    estimate = median(mean_count),
    lower = quantile(mean_count, 0.05),
    upper = quantile(mean_count, 0.95),
    .groups = "drop"
  )

interval_summary |>
  filter(Tech != 'Ceramic LSA') |>
  
  ggplot(aes(x = Tech, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.15
  ) +
  facet_wrap(~cluster_cat, scales = "free") +
  ylab("Radiocarbon Intensity") +
  xlab("Regional Aggregate") +
  theme_minimal()+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

interval_summary |>
  filter(Tech != 'Ceramic LSA') |>
  
  ggplot(aes(x = cluster_cat, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.15
  ) +
  facet_wrap(~Tech, scales = 'free') +
  ylab("Radiocarbon Intensity") +
  xlab("Regional Aggregate") +
  theme_minimal()+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r}
draws <- as_draws_df(fit_space_h4)


cluster_slopes <- draws |>
  select(
    .draw,
    starts_with("r_cluster_cat[")
  ) |>
  pivot_longer(
    cols = starts_with("r_cluster_cat["),
    names_to = "cluster",
    values_to = "cluster_deviation"
  ) |>
  mutate(
    cluster = 
      str_extract(cluster, "(?<=\\[)[^,]+"),
  ) |>
  select(.draw, cluster, cluster_deviation)

tech_slopes <- draws |>
  select(
    .draw,
    matches("^r_Tech\\[.*,time_ka\\]$")
  ) |>
  pivot_longer(
    cols = -.draw,
    names_to = "Tech",
    values_to = "tech_deviation"
  ) |>
  mutate(
    Tech = 
      str_extract(Tech, "(?<=\\[)[^,]+")
  ) |>
  select(.draw, Tech, tech_deviation)


fixed_slope <- draws |>
  select(.draw, b_time_ka)

slope_draws <- cluster_slopes |>
  left_join(
    tech_slopes,
    by = ".draw",
    relationship = "many-to-many"
  ) |>
  left_join(
    fixed_slope,
    by = ".draw"
  ) |>
  mutate(
    cluster_tech = paste(cluster, Tech, sep = "_"),
    cluster_slope =
      b_time_ka +
      cluster_deviation +
      tech_deviation
  )


slope_pairs <- combn(
  unique(slope_draws$cluster_tech),
  2,
  simplify = FALSE
)


slope_pair_comp <- tibble(
  combination_1 = character(),
  combination_2 = character(),
  estimate = numeric(),
  lower = numeric(),
  upper = numeric(),
  prob_1_greater = numeric(),
  rate_ratio = numeric(),
  rate_ratio_lower = numeric(),
  rate_ratio_upper = numeric()
)
for(i in slope_pairs){
  

    
    wide <- slope_draws |>
      filter(
        cluster_tech %in% i
      ) |>
      select(.draw, cluster_tech, cluster_slope) |>
      mutate(
        cluster_tech = as.character(cluster_tech)
      ) |>
      pivot_wider(
        id_cols = .draw,
        names_from = cluster_tech,
        values_from = cluster_slope
      )
    
    diff <- wide[,2] - wide[,3]
    
    tmp_tib <- tibble(
      combination_1 = i[1],
      combination_2 = i[2],
      estimate = median(diff[[1]]),
      lower = quantile(diff[[1]], 0.05),
      upper = quantile(diff[[1]], 0.95),
      prob_1_greater = mean(diff[[1]] > 0),
      rate_ratio = median(exp(diff[[1]])),
      rate_ratio_lower = quantile(exp(diff[[1]]), 0.05),
      rate_ratio_upper = quantile(exp(diff[[1]]), 0.95)
    )
    
    slope_pair_comp <- slope_pair_comp |>
      add_row(tmp_tib)
}
```

```{r}
# Track interior karoo rates across technologicl classifications
slope_pair_comp |>
  filter(str_detect(combination_1, 'Karoo') & 
           str_detect(combination_2, 'Karoo')) |>
  select(combination_1:combination_2, rate_ratio:rate_ratio_upper)


# Track southern coast rates across technologicl classifications
slope_pair_comp |>
  filter(str_detect(combination_1, 'Southern') & 
           str_detect(combination_2, 'Southern')) |>
  select(combination_1:combination_2, rate_ratio:rate_ratio_upper)

# Track southern coast rates across technologicl classifications
slope_pair_comp |>
  filter(str_detect(combination_1, 'East') & 
           str_detect(combination_2, 'East')) |>
  select(combination_1:combination_2, rate_ratio:rate_ratio_upper)
```

# Change Points

```{r}
time_mean <- mean(fit_df$calBP, na.rm = TRUE)

interval_grid <- crossing(
  intervals,
  cluster_cat = unique(fit_df$cluster_cat),
  Tech = unique(fit_df$Tech),
  calBP = seq(1000, 21000, by = 500)
) |>
  filter(
    calBP >= pmin(start, end),
    calBP <= pmax(start, end)
  ) |>
  mutate(
    time_ka = -(calBP - time_mean)/1000
  )

interval_pred <- posterior_epred(
  fit_space_h4,
  newdata = interval_grid,
  re_formula = NULL
)


pred_long <- interval_grid |>
  select(cluster_cat, Tech, calBP) |>
  bind_cols(
    as.data.frame(t(interval_pred))
  ) |>
  pivot_longer(
    cols = starts_with("V"),
    names_to = "draw",
    values_to = "mu"
  ) |>
  group_by(cluster_cat, Tech, calBP) |>
  summarize(Estimate = mean(mu),
            Q2.5 = quantile(mu, 0.025),
            Q97.5 = quantile(mu, 0.975))


# plot robberg distribution
ggplot(
  pred_long |> 
    filter(calBP > 7000,
           Tech == 'Robberg'),
  aes(
    x = calBP,
    y = Estimate,
    group = cluster_cat,
    colour = cluster_cat
  )
) +
  geom_ribbon(
    aes(
      ymin = Q2.5,
      ymax = Q97.5,
      fill = cluster_cat
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 1) +
  facet_wrap(~Tech, scales = 'free') +
  labs(
    x = "Calendar age (cal BP)",
    y = "Radiocarbon Intensity",
    colour = "Cluster",
    fill = "Cluster"
  ) +
  scale_x_reverse()+
  theme_classic() -> gg_robberg

# plot oakhurst distribution
ggplot(
  pred_long |> 
    filter(calBP > 5000,
           Tech == 'Oakhurst'),
  aes(
    x = calBP,
    y = Estimate,
    group = cluster_cat,
    colour = cluster_cat
  )
) +
  geom_ribbon(
    aes(
      ymin = Q2.5,
      ymax = Q97.5,
      fill = cluster_cat
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 1) +
  facet_wrap(~Tech, scales = 'free') +
  labs(
    x = "Calendar age (cal BP)",
    y = "Radiocarbon Intensity",
    colour = "Cluster",
    fill = "Cluster"
  ) +
  scale_x_reverse()+
  theme_classic() -> gg_oakhurst

# plot wilton distribution
ggplot(
  pred_long |> 
    filter((calBP > 4000) & (calBP < 12000),
           Tech == 'Wilton'),
  aes(
    x = calBP,
    y = Estimate,
    group = cluster_cat,
    colour = cluster_cat
  )
) +
  geom_ribbon(
    aes(
      ymin = Q2.5,
      ymax = Q97.5,
      fill = cluster_cat
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 1) +
  facet_wrap(~Tech, scales = 'free') +
  labs(
    x = "Calendar age (cal BP)",
    y = "Radiocarbon Intensity",
    colour = "Cluster",
    fill = "Cluster"
  ) +
  scale_x_reverse()+
  theme_classic() -> gg_wilton

# plot final LSA distribution
ggplot(
  pred_long |> 
    filter(calBP < 7000,
           Tech == 'Final LSA'),
  aes(
    x = calBP,
    y = Estimate,
    group = cluster_cat,
    colour = cluster_cat
  )
) +
  geom_ribbon(
    aes(
      ymin = Q2.5,
      ymax = Q97.5,
      fill = cluster_cat
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 1) +
  facet_wrap(~Tech, scales = 'free') +
  labs(
    x = "Calendar age (cal BP)",
    y = "Radiocarbon Intensity",
    colour = "Cluster",
    fill = "Cluster"
  ) +
  scale_x_reverse()+
  theme_classic() -> gg_finalLSA 
```

```{r}
ggarrange(
  gg_robberg, gg_oakhurst,
  gg_wilton, gg_finalLSA,
  
  common.legend = TRUE
)
```

```{r}
time_mean <- mean(fit_df$calBP, na.rm = TRUE)

interval_grid <- crossing(
  intervals,
  cluster_cat = unique(fit_df$cluster_cat),
  calBP = seq(1000, 21000, by = 500)
) |>
  filter(
    calBP >= pmin(start, end),
    calBP <= pmax(start, end)
  ) |>
  mutate(
    time_ka = -(calBP - time_mean)/1000
  )

interval_pred <- posterior_epred(
  fit_space_h3,
  newdata = interval_grid,
  re_formula = NULL
)


pred_long <- interval_grid |>
  select(period, cluster_cat, calBP) |>
  bind_cols(
    as.data.frame(t(interval_pred))
  ) |>
  pivot_longer(
    cols = starts_with("V"),
    names_to = "draw",
    values_to = "mu"
  ) |>
  group_by(cluster_cat, period, calBP) |>
  summarize(Estimate = mean(mu),
            Q2.5 = quantile(mu, 0.025),
            Q97.5 = quantile(mu, 0.975))


# plot robberg distribution
ggplot(
  pred_long |> 
    filter(period == 'Robberg'),
  aes(
    x = calBP,
    y = Estimate,
    group = cluster_cat,
    colour = cluster_cat
  )
) +
  geom_ribbon(
    aes(
      ymin = Q2.5,
      ymax = Q97.5,
      fill = cluster_cat
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 1) +
  labs(
    x = "Calendar age (cal BP)",
    y = "Radiocarbon Intensity",
    colour = "Cluster",
    fill = "Cluster"
  ) +
  scale_x_reverse()+
  theme_classic() -> gg_robberg

# plot oakhurst distribution
ggplot(
  pred_long |> 
    filter(period == 'Oakhurst'),
  aes(
    x = calBP,
    y = Estimate,
    group = cluster_cat,
    colour = cluster_cat
  )
) +
  geom_ribbon(
    aes(
      ymin = Q2.5,
      ymax = Q97.5,
      fill = cluster_cat
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 1) +
  labs(
    x = "Calendar age (cal BP)",
    y = "Radiocarbon Intensity",
    colour = "Cluster",
    fill = "Cluster"
  ) +
  scale_x_reverse()+
  theme_classic() -> gg_oakhurst

# plot wilton distribution
ggplot(
  pred_long |> 
    filter(period == 'Wilton'),
  aes(
    x = calBP,
    y = Estimate,
    group = cluster_cat,
    colour = cluster_cat
  )
) +
  geom_ribbon(
    aes(
      ymin = Q2.5,
      ymax = Q97.5,
      fill = cluster_cat
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 1) +
  labs(
    x = "Calendar age (cal BP)",
    y = "Radiocarbon Intensity",
    colour = "Cluster",
    fill = "Cluster"
  ) +
  scale_x_reverse()+
  theme_classic() -> gg_wilton

# plot final LSA distribution
ggplot(
  pred_long |> 
    filter(period == 'Final LSA'),
  aes(
    x = calBP,
    y = Estimate,
    group = cluster_cat,
    colour = cluster_cat
  )
) +
  geom_ribbon(
    aes(
      ymin = Q2.5,
      ymax = Q97.5,
      fill = cluster_cat
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 1) +
  labs(
    x = "Calendar age (cal BP)",
    y = "Radiocarbon Intensity",
    colour = "Cluster",
    fill = "Cluster"
  ) +
  scale_x_reverse()+
  theme_classic() -> gg_finalLSA 
```

```{r}
ggarrange(
  gg_robberg, gg_oakhurst,
  gg_wilton, gg_finalLSA,
  
  common.legend = TRUE
)
```

# Junks

```{r}
# compare at different points in time
prediction_grid <- expand_grid(
  calBP = seq(min_age, max_age, by = 1000),
  cluster_fct = unique(fit_df$cluster_fct)
) |>
  mutate(time_ka = -(calBP - mean(calBP, na.rm = TRUE))/1000)


pred <- fitted(
  fit_space_h2,
  newdata = prediction_grid,
  re_formula = NULL,
  summary = TRUE,
  allow_new_levels = TRUE
)

prediction_grid <- prediction_grid |>
  bind_cols(
    as.data.frame(pred)
  )



ggplot(
  prediction_grid |> filter(calBP > 10000),
  aes(
    x = calBP,
    y = Estimate,
    group = cluster_fct,
    colour = cluster_fct
  )
) +
  geom_ribbon(
    aes(
      ymin = Q2.5,
      ymax = Q97.5,
      fill = cluster_fct
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 1) +
  scale_x_reverse() +
  labs(
    x = "Calendar age (cal BP)",
    y = "Expected site occurrence",
    colour = "Cluster",
    fill = "Cluster"
  ) +
  theme_classic()


######################################################
######################################################

# compare at different points in time





prediction_grid <- expand_grid(
  calBP = seq(min_age, max_age, by = 1000),
  cluster_cat = unique(fit_df$cluster_cat)
) |>
  mutate(time_ka = -(calBP - time_mean)/1000)


pred <- fitted(
  fit_space_h3,
  newdata = prediction_grid,
  re_formula = NULL,
  summary = TRUE
)

prediction_grid <- prediction_grid |>
  bind_cols(
    as.data.frame(pred)
  )



ggplot(
  prediction_grid |> 
    filter(cluster_cat %in% c('Southern Coast', 'Interior-Karoo'),
           calBP > 8000),
  aes(
    x = calBP,
    y = Estimate,
    group = cluster_cat,
    colour = cluster_cat
  )
) +
  geom_ribbon(
    aes(
      ymin = Q2.5,
      ymax = Q97.5,
      fill = cluster_cat
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 1) +
  labs(
    x = "Calendar age (cal BP)",
    y = "Radiocarbon Intensity",
    colour = "Cluster",
    fill = "Cluster"
  ) +
  scale_x_reverse()+
  theme_classic()
```

```{r}
# Compare age ranges
intervals <- tribble(
  ~period, ~start, ~end,
  "Robberg", 13145, 19923,
  "Oakhurst",   8705, 12789,
  "Wilton",    4636,  7784,
  "Final Later Stone Age", 1318, 3898
) |>
  mutate(period = as.factor(period),
         period = fct_relevel(period, c('Robberg', 'Oakhurst', 'Wilton',
                                        'Final Later Stone Age')))

interval_grid <- crossing(
  intervals,
  cluster_fct = levels(fit_df$cluster_fct),
  calBP = seq(1000, 18000, by = 500)
) |>
  filter(
    calBP >= pmin(start, end),
    calBP <= pmax(start, end)
  ) |>
  mutate(
    time_ka = -(calBP - mean(calBP, na.rm = TRUE))/1000
  )

interval_pred <- posterior_epred(
  fit_space_h2,
  newdata = interval_grid,
  re_formula = NULL,
  allow_new_levels = TRUE
)


pred_long <- interval_grid |>
  select(period, cluster_fct, calBP) |>
  bind_cols(
    as.data.frame(t(interval_pred))
  ) |>
  pivot_longer(
    cols = starts_with("V"),
    names_to = "draw",
    values_to = "mu"
  )

# caculate expectations
interval_summary <- pred_long |>
  group_by(period, cluster_fct, draw) |>
  summarise(
    mean_count = mean(mu),
    .groups = "drop"
  ) |>
  group_by(period, cluster_fct) |>
  summarise(
    estimate = median(mean_count),
    lower = quantile(mean_count, 0.05),
    upper = quantile(mean_count, 0.95),
    .groups = "drop"
  )


interval_summary |>
  
  ggplot(aes(x = cluster_fct, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.15
  ) +
  facet_grid(~period, scales = "free_y") +
  ylab("Expected radiocarbon date values") +
  theme_minimal()+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
###################################################
###################################################
###################################################
time_mean <- mean(fit_df$calBP, na.rm = TRUE)

interval_grid <- crossing(
  intervals,
  cluster_cat = unique(fit_df$cluster_cat),
  calBP = seq(1000, 18000, by = 500)
) |>
  filter(
    calBP >= pmin(start, end),
    calBP <= pmax(start, end)
  ) |>
  mutate(
    time_ka = -(calBP - time_mean)/1000
  )

interval_pred <- posterior_epred(
  fit_space_h3,
  newdata = interval_grid,
  re_formula = NULL
)


pred_long <- interval_grid |>
  select(period, cluster_cat, calBP) |>
  bind_cols(
    as.data.frame(t(interval_pred))
  ) |>
  pivot_longer(
    cols = starts_with("V"),
    names_to = "draw",
    values_to = "mu"
  )

# caculate expectations
interval_summary <- pred_long |>
  group_by(period, cluster_cat, draw) |>
  mutate(period = as.factor(period)) |> 
  summarise(
    mean_count = mean(mu),
    .groups = "drop"
  ) |>
  group_by(period, cluster_cat) |>
  summarise(
    estimate = median(mean_count),
    lower = quantile(mean_count, 0.05),
    upper = quantile(mean_count, 0.95),
    .groups = "drop"
  )

interval_summary |>
      
  ggplot(aes(x = cluster_cat, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.15
  ) +
  facet_wrap(~period, scales = "free_y") +
  ylab("Radiocarbon Intensity") +
  xlab("Regional Aggregate") +
  theme_minimal()+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) 
```

### Interval rate ratio

```{r}
interval_draws <- pred_long |>
  group_by(period, cluster_fct, draw) |>
  summarise(
    mean_count = mean(mu),
    .groups = "drop"
  )

clusters <- unique(interval_draws$cluster_fct)

cluster_pairs <- combn(
  clusters,
  2,
  simplify = FALSE
)

interval_rate_ratios <- map_dfr(
  cluster_pairs,
  \(pair) {
    
    a <- interval_draws |>
      filter(cluster_fct == pair[1]) |>
      select(period, draw, count_a = mean_count)
    
    b <- interval_draws |>
      filter(cluster_fct == pair[2]) |>
      select(period, draw, count_b = mean_count)
    
    left_join(a, b, by = c("period", "draw")) |>
      mutate(
        cluster_1 = pair[1],
        cluster_2 = pair[2],
        rate_ratio = count_a / count_b
      )
  }
)


interval_rate_ratio_summary <- interval_rate_ratios |>
  group_by(period, cluster_1, cluster_2) |>
  summarise(
    estimate = median(rate_ratio),
    lower = quantile(rate_ratio, 0.025),
    upper = quantile(rate_ratio, 0.975),
    prob_greater_1 = mean(rate_ratio > 1),
    .groups = "drop"
  )

interval_rate_ratio_summary |>
  filter(prob_greater_1 > 0.9 | prob_greater_1 < 0.1)


####################################################################
####################################################################
####################################################################
interval_draws <- pred_long |>
  group_by(period, cluster_cat, draw) |>
  summarise(
    mean_count = mean(mu),
    .groups = "drop"
  )

clusters <- unique(interval_draws$cluster_cat)

cluster_pairs <- combn(
  clusters,
  2,
  simplify = FALSE
)

interval_rate_ratios <- map_dfr(
  cluster_pairs,
  \(pair) {
    
    a <- interval_draws |>
      filter(cluster_cat == pair[1]) |>
      select(period, draw, count_a = mean_count)
    
    b <- interval_draws |>
      filter(cluster_cat == pair[2]) |>
      select(period, draw, count_b = mean_count)
    
    left_join(a, b, by = c("period", "draw")) |>
      mutate(
        cluster_1 = pair[1],
        cluster_2 = pair[2],
        rate_ratio = count_a / count_b
      )
  }
)


interval_rate_ratio_summary <- interval_rate_ratios |>
  group_by(period, cluster_1, cluster_2) |>
  summarise(
    estimate = median(rate_ratio),
    lower = quantile(rate_ratio, 0.05),
    upper = quantile(rate_ratio, 0.95),
    prob_greater_1 = mean(rate_ratio > 1),
    .groups = "drop"
  )

interval_rate_ratio_summary |>
  filter(prob_greater_1 > 0.90 | prob_greater_1 < 0.10) |>
  mutate(estimate = estimate,
         lower = lower,
         upper = upper)
```

### Within cluster change

```{r}
# temporal change within each cluster
temporal_change <- interval_draws |>
  #filter(period %in% c("18–12 ka", "12–8 ka", "8-4 ka", "4-2 ka")) |>
  select(cluster_fct, period, draw, mean_count) |>
  pivot_wider(
    names_from = period,
    values_from = mean_count
  ) |>
  mutate(
    rate_ratio_Rob_Oak = `12–8 ka` / `18–12 ka`,
    rate_ratio_Oak_Wilt = `8–4 ka` / `12–8 ka`,
    rate_ratio_Wilt_Post = `4–2 ka` / `8–4 ka`
  ) 



temporal_change |>
  pivot_longer('12–8 ka':'rate_ratio_Wilt_Post', 
               names_to = 'Rates', values_to = 'Pred') |>
  group_by(Rates) |>
  summarise(
    estimate = round(median(Pred), 3),
    lower = round(quantile(Pred, 0.025), 3),
    upper = round(quantile(Pred, 0.975), 3),
    .groups = "drop"
  )


###########################################################
###########################################################
###########################################################

# temporal change within each cluster
temporal_change <- interval_draws |>
  #filter(period %in% c("18–12 ka", "12–8 ka", "8-4 ka", "4-2 ka")) |>
  select(cluster_cat, period, draw, mean_count) |>
  pivot_wider(
    names_from = period,
    values_from = mean_count
  ) |>
  mutate(
    rate_ratio_Rob_Oak = `12–8 ka` / `18–12 ka`,
    rate_ratio_Oak_Wilt = `8–4 ka` / `12–8 ka`,
    rate_ratio_Wilt_Post = `4–2 ka` / `8–4 ka`
  ) 



temporal_change |>
  pivot_longer('12–8 ka':'rate_ratio_Wilt_Post', 
               names_to = 'Rates', values_to = 'Pred') |>
  group_by(Rates) |>
  summarise(
    estimate = round(median(Pred), 3),
    lower = round(quantile(Pred, 0.025), 3),
    upper = round(quantile(Pred, 0.975), 3),
    .groups = "drop"
  )

```

# Biome modeling

```{r}
# combine clustered data from SASSI to the SARD database
df_tmp <- za_holocene |>
  st_drop_geometry() |>
  ungroup() |>
  rename(Site_Type = 'Site Type') |>
  
  mutate(time_s = as.numeric(scale(calBP)),
         time_red = calBP/1000,
         time_ka = (calBP - mean(calBP, na.rm = TRUE))/1000,,
         
         Site_Type = as.factor(Site_Type),
         
         iter = iter)
```

```{r}

fit_df |>
  ggplot(aes(x = calBP, y = site_count, fill = site_count))+
  geom_bar(stat = "identity") +
  facet_wrap(~Biome, scale = "free") +
  scale_fill_gradient(low = "darkorange", high = "darkblue") +
  xlab("Calibrated Years BP")+
  theme_minimal(base_size = 8)+
  theme(axis.text.x = element_text(angle = 45,  hjust = 1))


fit_df |>
  ggplot(aes(x = calBP, y = site_count, fill = site_count))+
  geom_bar(stat = "identity") +
  facet_wrap(~Site_Type, scale = "free") +
  scale_fill_gradient(low = "darkorange", high = "darkblue") +
  xlab("Calibrated Years BP")+
  theme_minimal(base_size = 8)+
  theme(axis.text.x = element_text(angle = 45,  hjust = 1))
```

## Bayesian Modeling

```{r}
# this model shows cluster show almost no variation after accounting for mean site count
seed <- seed + 1
fit_time <- brm(
  site_count ~
    1 + time_ka,
  
  family = negbinomial(),
  
  data = fit_df,
  
  prior = c(
    prior(normal(0, 2), class = "Intercept")
  ),
  
  chains = 1,
  iter = 4000,
  warmup = 2000,
  
  control = list(
    adapt_delta = 0.98
  ),
  
  cores = 4,
  threads = threading(8),
  backend = 'cmdstanr',
  seed = seed
)

#######################################################

seed <- seed + 1
fit_space_biome <- brm(
  site_count ~
    1 + time_ka + (0 + time_ka || Biome),
  
  family = negbinomial(),
  
  data = fit_df,
  
  prior = c(
    prior(normal(0, 2), class = "Intercept")
  ),
  
  chains = 1,
  iter = 4000,
  warmup = 2000,
  
  control = list(
    adapt_delta = 0.95
  ),
  
  cores = 4,
  threads = threading(8),
  backend = 'cmdstanr',
  seed = seed
)

############################################################
seed <- seed + 1
fit_space_site_type <- brm(
  site_count ~
    1 + time_ka + (0 + time_ka || Site_Type),
  
  family = negbinomial(),
  
  data = fit_df,
  
  prior = c(
    prior(normal(0, 2), class = "Intercept")
  ),
  
  chains = 1,
  iter = 4000,
  warmup = 2000,
  
  control = list(
    adapt_delta = 0.95
  ),
  
  cores = 4,
  threads = threading(8),
  backend = 'cmdstanr',
  seed = seed
)
```

```{r}
# compute posterior slope estimates for each cluster
cluster_slope_draws <- spread_draws(
  fit_space_biome,
  r_Biome[Biome, time_ka]
)


cluster_slope_summary <- cluster_slope_draws |>
  group_by(Biome) |>
  summarise(
    estimate = median(r_Biome),
    lower = quantile(r_Biome, 0.05),
    upper = quantile(r_Biome, 0.95),
    .groups = "drop"
  )




draws <- as_draws_df(fit_space_biome)

slope_draws <- draws |>
  select(
    .draw,
    b_time_ka,
    starts_with("r_Biome[")
    ) |>
  pivot_longer(
    cols = starts_with("r_Biome["),
    names_to = "biome",
    values_to = "slope_deviation"
    ) |>
  mutate(
    biome_slope = b_time_ka + slope_deviation,
    biome = str_extract(biome, "(?<=\\[)[^,]+")
    ) |>
  group_by(biome) |>
  summarise(
    slope = median(biome_slope),
    lower = quantile(biome_slope, 0.05),
    upper = quantile(biome_slope, 0.95),
    
    rate_ratio_1ka = exp(slope),
    rate_lower = exp(lower),
    rate_upper = exp(upper),
    
    .groups = "drop"
    )

slope_draws

##############################################################
##############################################################
##############################################################

# compute posterior slope estimates for each cluster
cluster_slope_draws <- spread_draws(
  fit_space_site_type,
  r_Site_Type[Site_Type, time_ka]
)


cluster_slope_summary <- cluster_slope_draws |>
  group_by(Site_Type) |>
  summarise(
    estimate = median(r_Site_Type),
    lower = quantile(r_Site_Type, 0.025),
    upper = quantile(r_Site_Type, 0.975),
    .groups = "drop"
  )




draws <- as_draws_df(fit_space_site_type)

slope_draws <- draws |>
  select(
    .draw,
    b_time_ka,
    starts_with("r_Site_Type[")
  ) |>
  pivot_longer(
    cols = starts_with("r_Site_Type["),
    names_to = "site_type",
    values_to = "slope_deviation"
  ) |>
  mutate(
    site_type_slope = b_time_ka + slope_deviation,
    site_type = str_extract(site_type, "(?<=\\[)[^,]+")
  ) |>
  group_by(site_type) |>
  summarise(
    slope = median(site_type_slope),
    lower = quantile(site_type_slope, 0.025),
    upper = quantile(site_type_slope, 0.975),
    
    rate_ratio_1ka = exp(slope),
    rate_lower = exp(lower),
    rate_upper = exp(upper),
    
    .groups = "drop"
  )

slope_draws
```

```{r}
# slope comparisons
slope_draws <- as.data.frame(
  coef(fit_space_biome, summary = FALSE)$Biome
)

slope_pairs <- combn(
  colnames(slope_draws),
  2,
  simplify = FALSE
)

pairwise_slopes <- map_dfr(
  slope_pairs,
  \(x) {
    
    diff <- slope_draws[[x[1]]] - slope_draws[[x[2]]]
    
    tibble(
      Biome_1 = x[1],
      Biome_2 = x[2],
      difference = diff,
      estimate = median(diff),
      lower = quantile(diff, 0.025),
      upper = quantile(diff, 0.975),
      prob_biome1_greater = mean(diff > 0)
    )
  }
)

# shows that no slopes are actually significantly different
pairwise_slopes |> 
  distinct(lower, upper, .keep_all = TRUE) |>
  filter((lower < 0 & upper < 0) | (lower > 0 & upper > 0),
         str_detect(Biome_1, 'ka') & str_detect(Biome_2, 'ka'))
  
###############################################################################
###############################################################################
###############################################################################
  
  # slope comparisons
  slope_draws <- as.data.frame(
    coef(fit_space_site_type, summary = FALSE)$Site_Type
  )

  slope_pairs <- combn(
    colnames(slope_draws),
    2,
    simplify = FALSE
  )
  
  pairwise_slopes <- map_dfr(
    slope_pairs,
    \(x) {
      
      diff <- slope_draws[[x[1]]] - slope_draws[[x[2]]]
      
      tibble(
        siteType_1 = x[1],
        siteType_2 = x[2],
        difference = diff,
        estimate = median(diff),
        lower = quantile(diff, 0.05),
        upper = quantile(diff, 0.95),
        prob_siteType1_greater = mean(diff > 0)
      )
    }
  )
  
  # shows that no slopes are actually significantly different
  pairwise_slopes |> 
    distinct(lower, upper, .keep_all = TRUE) |>
  filter(prob_siteType1_greater <= 0.1 | prob_siteType1_greater >= 0.9,
         str_detect(siteType_1, 'ka') & str_detect(siteType_2, 'ka'))
```

```{r}
# compare at different points in time
prediction_grid <- expand_grid(
  calBP = seq(min_age, max_age, by = 1000),
  Biome = unique(fit_df$Biome)
) |>
  mutate(time_ka = (calBP - mean(calBP, na.rm = TRUE))/1000)


pred <- fitted(
  fit_space_biome,
  newdata = prediction_grid,
  re_formula = NULL,
  summary = TRUE,
  allow_new_levels = TRUE
)

prediction_grid <- prediction_grid |>
  bind_cols(
    as.data.frame(pred)
  )


#prediction_grid |> filter(calBP == 12000,cluster_fct %in% c("3", "5") )

ggplot(
  prediction_grid |> filter(calBP > 12000, !is.na(Biome)),
  aes(
    x = calBP,
    y = Estimate,
    group = Biome,
    colour = Biome
  )
) +
  geom_ribbon(
    aes(
      ymin = Q2.5,
      ymax = Q97.5,
      fill = Biome
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 1) +
  scale_x_reverse() +
  labs(
    x = "Calendar age (cal BP)",
    y = "Expected site occurrence",
    colour = "Cluster",
    fill = "Cluster"
  ) +
  theme_classic()


######################################################
######################################################

# compare at different points in time
prediction_grid <- expand_grid(
  calBP = seq(min_age, max_age, by = 1000),
  cluster_cat = unique(fit_df$cluster_cat)
) |>
  mutate(time_ka = (calBP - mean(calBP, na.rm = TRUE))/1000)


pred <- fitted(
  fit_space_h3,
  newdata = prediction_grid,
  re_formula = NULL,
  summary = TRUE
)

prediction_grid <- prediction_grid |>
  bind_cols(
    as.data.frame(pred)
  )


prediction_grid |>
  filter(
    calBP == 12000,
    cluster_cat %in% c("Southern Coastal", "Interior")
  )



ggplot(
  prediction_grid |> 
    filter(cluster_cat %in% c('Southern Coast', 'Interior-Karoo', 'Northeast')),
  aes(
    x = calBP,
    y = Estimate,
    group = cluster_cat,
    colour = cluster_cat
  )
) +
  geom_ribbon(
    aes(
      ymin = Q2.5,
      ymax = Q97.5,
      fill = cluster_cat
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 1) +
  scale_x_reverse() +
  labs(
    x = "Calendar age (cal BP)",
    y = "Radiocarbon Intensity",
    colour = "Cluster",
    fill = "Cluster"
  ) +
  theme_classic()



ggplot(
  prediction_grid |> 
    filter(cluster_cat %in% c('Southern Coast', 'Interior-Karoo', 'Northeast'),
           calBP > 9000),
  aes(
    x = calBP,
    y = Estimate,
    group = cluster_cat,
    colour = cluster_cat
  )
) +
  geom_ribbon(
    aes(
      ymin = Q2.5,
      ymax = Q97.5,
      fill = cluster_cat
    ),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 1) +
  scale_x_reverse() +
  labs(
    x = "Calendar age (cal BP)",
    y = "Radiocarbon Intensity",
    colour = "Cluster",
    fill = "Cluster"
  ) +
  theme_classic()
```

```{r}
# COompare age ranges
intervals <- tribble(
  ~period, ~start, ~end,
  "18–12 ka", 12000, 18000,
  "12–8 ka", 8000, 12000,
  "8–4 ka",   4000, 8000,
  "4–2 ka",    2000,  4000
)

interval_grid <- crossing(
  intervals,
  Biome = unique(fit_df$Biome),
  calBP = seq(2000, 18000, by = 500)
) |>
  filter(
    calBP >= pmin(start, end),
    calBP <= pmax(start, end)
  ) |>
  mutate(
    time_ka = (calBP - mean(calBP, na.rm = TRUE))/1000
  )

interval_pred <- posterior_epred(
  fit_space_biome,
  newdata = interval_grid,
  re_formula = NULL,
  allow_new_levels = TRUE
)


pred_long <- interval_grid |>
  select(period, Biome, calBP) |>
  bind_cols(
    as.data.frame(t(interval_pred))
  ) |>
  pivot_longer(
    cols = starts_with("V"),
    names_to = "draw",
    values_to = "mu"
  )

# caculate expectations
interval_summary <- pred_long |>
  group_by(period, Biome, draw) |>
  summarise(
    mean_count = mean(mu),
    .groups = "drop"
  ) |>
  group_by(period, Biome) |>
  summarise(
    estimate = median(mean_count),
    lower = quantile(mean_count, 0.05),
    upper = quantile(mean_count, 0.95),
    .groups = "drop"
  ) |>
  filter(!is.na(Biome))


interval_summary |>
  mutate(period = fct_relevel(period,
                              "18–12 ka", "12–8 ka", "8–4 ka", "4–2 ka")) |>
  
  ggplot(aes(x = Biome, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.15
  ) +
  facet_wrap(~period, scales = "free_y") +
  ylab("Expected radiocarbon date values") +
  theme_minimal()+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
###################################################
###################################################
###################################################
interval_grid <- crossing(
  intervals,
  cluster_cat = unique(fit_df$cluster_cat),
  calBP = seq(2000, 18000, by = 500)
) |>
  filter(
    calBP >= pmin(start, end),
    calBP <= pmax(start, end)
  ) |>
  mutate(
    time_ka = (calBP - mean(calBP, na.rm = TRUE))/1000
  )

interval_pred <- posterior_epred(
  fit_space_h3,
  newdata = interval_grid,
  re_formula = NULL
)


pred_long <- interval_grid |>
  select(period, cluster_cat, calBP) |>
  bind_cols(
    as.data.frame(t(interval_pred))
  ) |>
  pivot_longer(
    cols = starts_with("V"),
    names_to = "draw",
    values_to = "mu"
  )

# caculate expectations
interval_summary <- pred_long |>
  group_by(period, cluster_cat, draw) |>
  mutate(period = as.factor(period)) |> 
  summarise(
    mean_count = mean(mu),
    .groups = "drop"
  ) |>
  group_by(period, cluster_cat) |>
  summarise(
    estimate = median(mean_count),
    lower = quantile(mean_count, 0.025),
    upper = quantile(mean_count, 0.975),
    .groups = "drop"
  )

interval_summary |>
  mutate(period = fct_relevel(period,
                              "18–12 ka", "12–8 ka", "8–4 ka", "4–2 ka")) |>
      
  ggplot(aes(x = cluster_cat, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    width = 0.15
  ) +
  facet_wrap(~period, scales = "free_y") +
  ylab("Radiocarbon Intensity") +
  theme_minimal()+
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) 
```

### Interval rate ratio

```{r}
interval_draws <- pred_long |>
  group_by(period, Biome, draw) |>
  summarise(
    mean_count = mean(mu),
    .groups = "drop"
  ) |>
  filter(!is.na(Biome))

biome <- unique(interval_draws$Biome)

biome_pairs <- combn(
  biome,
  2,
  simplify = FALSE
)

interval_rate_ratios <- map_dfr(
  biome_pairs,
  \(pair) {
    
    a <- interval_draws |>
      filter(Biome == pair[1]) |>
      select(period, draw, count_a = mean_count)
    
    b <- interval_draws |>
      filter(Biome == pair[2]) |>
      select(period, draw, count_b = mean_count)
    
    left_join(a, b, by = c("period", "draw")) |>
      mutate(
        biome_1 = pair[1],
        biome_2 = pair[2],
        rate_ratio = count_a / count_b
      )
  }
)


interval_rate_ratio_summary <- interval_rate_ratios |>
  group_by(period, biome_1, biome_2) |>
  summarise(
    estimate = median(rate_ratio),
    lower = quantile(rate_ratio, 0.05),
    upper = quantile(rate_ratio, 0.95),
    prob_greater_1 = mean(rate_ratio > 1),
    .groups = "drop"
  )

interval_rate_ratio_summary |>
  filter(prob_greater_1 > 0.9 | prob_greater_1 < 0.1)


####################################################################
####################################################################
####################################################################
interval_draws <- pred_long |>
  group_by(period, cluster_cat, draw) |>
  summarise(
    mean_count = mean(mu),
    .groups = "drop"
  )

clusters <- unique(interval_draws$cluster_cat)

cluster_pairs <- combn(
  clusters,
  2,
  simplify = FALSE
)

interval_rate_ratios <- map_dfr(
  cluster_pairs,
  \(pair) {
    
    a <- interval_draws |>
      filter(cluster_cat == pair[1]) |>
      select(period, draw, count_a = mean_count)
    
    b <- interval_draws |>
      filter(cluster_cat == pair[2]) |>
      select(period, draw, count_b = mean_count)
    
    left_join(a, b, by = c("period", "draw")) |>
      mutate(
        cluster_1 = pair[1],
        cluster_2 = pair[2],
        rate_ratio = count_a / count_b
      )
  }
)


interval_rate_ratio_summary <- interval_rate_ratios |>
  group_by(period, cluster_1, cluster_2) |>
  summarise(
    estimate = median(rate_ratio),
    lower = quantile(rate_ratio, 0.05),
    upper = quantile(rate_ratio, 0.95),
    prob_greater_1 = mean(rate_ratio > 1),
    .groups = "drop"
  )

interval_rate_ratio_summary |>
  filter(prob_greater_1 > 0.90 | prob_greater_1 < 0.10)
```

# Spatiotemporal

```{r}
fit_df <- za_holocene |>
  rename(site_count = count,
         Site_Type = `Site Type`) |>
  
  mutate(time_red = calBP/1000,
         time_ka = (calBP - mean(calBP, na.rm = TRUE))/1000,
         time_s = as.numeric(scale(calBP, center = TRUE, scale = TRUE)),
         
         lat_s = as.numeric(scale(y), center = TRUE, scale = TRUE),
         lon_s = as.numeric(scale(x), center = TRUE, scale = TRUE),
         
         Site_Type = as.factor(Site_Type),
         
         iter = iter) |>
  filter(!is.na(lat_s) & !is.na(lon_s))



```

```{r}
get_ig_params <- function(min_dist, max_dist, tail_prob = 0.01) {
  # Solves for inverse-gamma(shape, rate) such that:
  #   P(X < min_dist) = tail_prob
  #   P(X > max_dist) = tail_prob
  # This bounds the GP lengthscale prior between the smallest meaningful
  # distance in your data and the largest, avoiding boundary collapse.
  
  target <- function(par) {
    shape <- exp(par[1])  # exponentiate to keep search unconstrained but shape>0
    rate  <- exp(par[2])
    
    p_low  <- pinvgamma(min_dist, shape, rate)
    p_high <- 1 - pinvgamma(max_dist, shape, rate)
    
    # squared error against target tail probabilities
    (p_low - tail_prob)^2 + (p_high - tail_prob)^2
  }
  
  # inverse-gamma CDF via base R's gamma CDF (since 1/X ~ Gamma(shape, rate))
  pinvgamma <- function(x, shape, rate) {
    1 - pgamma(1 / x, shape = shape, rate = rate)
  }
  environment(target) <- environment()  # ensure pinvgamma is visible inside target
  
  # reasonable starting guess: shape=2, rate=min_dist (rough heuristic)
  init <- log(c(2, min_dist))
  
  opt <- optim(
    par = init,
    fn = target,
    method = "Nelder-Mead",
    control = list(reltol = 1e-12, maxit = 5000)
  )
  
  shape <- exp(opt$par[1])
  rate  <- exp(opt$par[2])
  
  # quick diagnostic printout
  cat(sprintf(
    "shape = %.4f, rate = %.4f\nP(X < %.4f) = %.4f\nP(X > %.4f) = %.4f\n",
    shape, rate,
    min_dist, pinvgamma(min_dist, shape, rate),
    max_dist, pinvgamma(max_dist, shape, rate)
  ))
  
  list(shape = shape, rate = rate)
}
```

```{r}
# find min/max pairwise distances in your standardized coords first
d <- as.matrix(dist(fit_df[, c("lon_s","lat_s")] |> 
                      st_drop_geometry() |> 
                      distinct()))
diag(d) <- NA
min_d <- min(d, na.rm=TRUE); max_d <- max(d, na.rm=TRUE)

ig <- get_ig_params(min_d, max_d)
```

```{r}

#seed <- seed + 1
seed <- 28
fit_spatiotemporal <- brm(
  site_count ~
    1 +
    time_s +
    gp(lon_s, lat_s),
  
  family = negbinomial(),
  
  data = fit_df,
  
  prior = c(
    prior(normal(0, 2), class = "Intercept"),
    prior(exponential(2), class = "sdgp"),
    prior(lognormal(log(1.5), 0.5), class = "lscale")
  ),
  
  chains = 2,
  iter = 4000,
  warmup = 2000,
  
  control = list(
    adapt_delta = 0.95
  ),
  
  cores = 4,
  threads = threading(8),
  backend = 'cmdstanr',
  seed = seed
)

#seed <- seed + 1
seed <- 29
fit_spatiotemporal2 <- brm(
  site_count ~
    1 +
    time_s +
    gp(lon_s, lat_s) +
    gp(lon_s, lat_s, by = time_s),
  
  family = negbinomial(),
  
  data = fit_df,
  
  prior = c(
    prior(normal(0, 2), class = "Intercept"),
    
    prior(exponential(2), class = "sdgp", coef = "gplon_slat_s"),
    prior(exponential(2), class = "sdgp", coef = "gplon_slat_stime_s"),
    
    prior(lognormal(log(1), 0.4), class = "lscale", coef = "gplon_slat_s"),
    prior(lognormal(log(1.5), 0.4), class = "lscale", coef = "gplon_slat_stime_s")
  ),
  
  chains = 1,
  iter = 4000,
  warmup = 2000,
  
  control = list(
    adapt_delta = 0.95
  ),
  
  cores = 4,
  threads = threading(8),
  backend = 'cmdstanr',
  seed = seed
)

####################
seed <- 30
fit_spatiotemporal3 <- brm(
  site_count ~
    1 +
    gp(lon_s, lat_s, k = 20, c = 5/4)+
    gp(time_s, k = 20, c = 5/4),
  
  family = negbinomial(),
  
  data = fit_df,
  
  prior = c(
    prior(normal(0, 2), class = "Intercept"),
    
    prior(exponential(2), class = "sdgp", coef = "gplon_slat_s"),
    prior(exponential(2), class = "sdgp", coef = "gptime_s"),
    
    prior(inv_gamma(0.51, 0.00059), class = "lscale", coef = "gplon_slat_s")
  ),
  
  chains = 1,
  iter = 4000,
  warmup = 2000,
  
  control = list(
    adapt_delta = 0.95
  ),
  
  cores = 4,
  threads = threading(8),
  backend = 'cmdstanr',
  seed = seed
)

```

```{r}
lon_center <- mean(fit_df$x)
lon_scale  <- sd(fit_df$x)

lat_center <- mean(fit_df$y)
lat_scale  <- sd(fit_df$y)

time_center <- mean(fit_df$calBP)
time_scale  <- sd(fit_df$calBP)


prediction_grid <- expand_grid(
  lon_s = seq(min(fit_df$lon_s), max(fit_df$lon_s),
              length.out = 20),
  lat_s = seq(min(fit_df$lat_s), max(fit_df$lat_s),
              length.out = 20),
  time_s = seq(min(fit_df$time_s), max(fit_df$time_s),
               length.out = 10),
) |>
  mutate(
    lon = (lon_s * lon_scale) + lon_center,
    lat = (lat_s * lat_scale) + lat_center,
    
    calBP = time_s * time_scale + time_center
  )

prediction_sf <- st_as_sf(
  prediction_grid,
  coords = c("lon", "lat"),
  crs = 4326,
  remove = FALSE)

xy <- st_coordinates(prediction_sf)

prediction_grid <- prediction_sf |>
  st_drop_geometry() |>
  mutate(
    x = xy[, 1],
    y = xy[, 2]
  )

post_pred <- posterior_epred(
  fit_spatiotemporal,
  newdata = prediction_grid,
  re_formula = NA
)

prediction_grid <- prediction_grid |>
  mutate(
    estimate = apply(post_pred, 2, median),
    lower = apply(post_pred, 2, quantile, probs = 0.05),
    upper = apply(post_pred, 2, quantile, probs = 0.95)
  )
```

```{r}
southern_africa_utm <- st_transform(
  africa_map,
  crs = 32735   # replace with the UTM CRS you're using
)


ggplot() +
  
  geom_raster(data = prediction_grid |> 
                filter(calBP == 2000),
              aes(x = lon, y = lat, fill = estimate)) +
  
  geom_sf(
    data = southern_africa_utm,
    fill = NA,
    color = "black",
    linewidth = 0.4
  )  +
  
  scale_fill_viridis_c() +
  coord_sf(expand = FALSE) +
  theme_minimal()

ggplot() +
  geom_sf(
    data = southern_africa_utm,
    fill = NA,
    color = "black",
    linewidth = 0.4
  ) +
  geom_sf(data = za_holocene, aes(color = `Biome`)) +
  scale_fill_viridis_c()

ggplot() +
  geom_sf(data = southern_africa_utm, fill = "grey90", color = "grey40")+
  coord_sf(xlim = c(min(za_holocene$x), max(za_holocene$x)), ylim = c(min(za_holocene$y), max(za_holocene$y)), expand = FALSE) +
  theme_minimal()
```

## Modeling by time

```{r}
# Spatial extent of your data
lon_range <- range(fit_df$x, na.rm = TRUE)
lat_range <- range(fit_df$y, na.rm = TRUE)

# Create a dense regular grid
pred_grid <- expand_grid(
  lon = seq(lon_range[1], lon_range[2], length.out = 50),
  lat = seq(lat_range[1], lat_range[2], length.out = 50)
)

lon_mean <- mean(fit_df$x)
lon_sd  <- sd(fit_df$x)

lat_mean <- mean(fit_df$y)
lat_sd  <- sd(fit_df$y)

time_mean <- mean(fit_df$calBP)
time_sd  <- sd(fit_df$calBP)

pred_grid <- pred_grid |>
  mutate(
    lon_s = (lon - lon_mean) / lon_sd,
    lat_s = (lat - lat_mean) / lat_sd,
    
    chunk = ceiling(row_number() / 1000)
  )

times <- c(
  20000, 18000, 16000, 14000, 12000,
  10000, 8000, 6000, 4000, 2000
)

pred_grid <- crossing(
  pred_grid,
  calBP = times
) |>
  mutate(
    time_s = (calBP - time_mean) / time_sd
  )



prediction_sf <- st_as_sf(
  pred_grid,
  coords = c("lon", "lat"),
  crs = 4326,
  remove = FALSE)

xy <- st_coordinates(prediction_sf)

pred_grid <- prediction_sf |>
  st_drop_geometry() |>
  mutate(
    lon = xy[, 1],
    lat = xy[, 2]
  )


pred_chunks <- lapply(
  unique(pred_grid$chunk),
  function(i) {
    
    message("Predicting chunk ", i, " of ",
            max(pred_grid$chunk))
    
    nd <- pred_grid |>
      filter(chunk == i)
    
    epred <- posterior_epred(
      fit_spatiotemporal2,
      newdata = nd,
      re_formula = NA,
      ndraws = 500
    )
    
    nd |>
      mutate(
        estimate = apply(epred, 2, median),
        lower = apply(epred, 2, quantile, probs = 0.05),
        upper = apply(epred, 2, quantile, probs = 0.95)
      )
  }
)

pred_summary <- bind_rows(pred_chunks)

```

```{r}

southern_africa <- ne_countries(
  scale = "medium",
  returnclass = "sf"
) |>
  filter(
    admin %in% c(
      "South Africa",
      "Namibia",
      "Botswana",
      "Zimbabwe",
      "Mozambique",
      "Eswatini",
      "Lesotho",
      "Zambia",
      "Malawi"
    )
  ) |>
  st_union()

pred_sf <- st_as_sf(
  pred_summary,
  coords = c("lon", "lat"),
  crs = 4326
)


pred_sf <- pred_sf[
  st_within(pred_sf, southern_africa, sparse = FALSE),
]

```

```{r}
ggplot() +
  
  geom_raster(
    data = pred_summary,
    aes(
      x = lon,
      y = lat,
      fill = estimate
    )
  ) +
  
  geom_sf(
    data = southern_africa,
    fill = NA,
    color = "black",
    linewidth = 0.4
  ) +
  
  coord_sf(
    xlim = range(pred_summary$lon),
    ylim = range(pred_summary$lat),
    expand = FALSE
  ) +
  
  scale_fill_viridis_c(
    name = "Predicted\nsite intensity"
  ) +
  
  theme_minimal()

ggplot(pred_summary) +
  geom_raster(
    aes(
      x = lon,
      y = lat,
      fill = estimate
    )
  ) +
  scale_fill_viridis_c(
    trans = scales::log10_trans(),
    name = "Predicted\nsite intensity"
  ) +
  coord_fixed() +
  theme_minimal()
```

# Junk

```{r}
pred_summary <- pred_long |>
  group_by(lon, lat, calBP) |>
  summarise(
    estimate = median(radiocarbon_rate),
    lower = quantile(radiocarbon_rate, 0.05),
    upper = quantile(radiocarbon_rate, 0.95),
    .groups = "drop"
  ) |>
  filter(!is.na(lon))


pred_sf <- st_as_sf(
  pred_summary,
  coords = c("lon", "lat"),
  crs = 32734
) |>
  st_transform(
    crs = 4326
  )

pred_coord <- pred_sf |>
  mutate(
    lon = st_coordinates(pred_sf)[,1],
    lat = st_coordinates(pred_sf)[,2]
  ) 


ggplot()+
  geom_sf(data = pred_coord, aes(x = lon, y = lat))


ggplot() +
  geom_sf(data = africa_map, fill = "grey90", color = "grey40") +
  
  geom_raster(data = pred_coord, aes(x = lon, y = lat, fill = estimate)) 
  
  coord_sf() +
  scale_fill_viridis_c() +
  theme_minimal()
```

```{r}
interp_surface <- interp(x = pred_coord$lon, y = pred_coord$lat, 
                        z = pred_coord$estimate, duplicate = 'mean')

surface_df <- expand.grid(
  lon = interp_surface$x,
  lat = interp_surface$y
) |>
  mutate(estimate = as.vector(interp_surface$z))

ggplot() +
  geom_sf(data = africa_map, fill = "grey90", color = "grey40") +
  geom_raster(
    data = surface_df,
    aes(
      x = lon,
      y = lat,
      fill = estimate
    )
  ) +
  coord_sf()





# 3. Convert the interpolated matrix back to a data frame for ggplot
continuous_surface <- expand.grid(lon = spline_interp$x, lat = spline_interp$y)
continuous_surface$value <- as.vector(spline_interp$z)
continuous_surface <- na.omit(continuous_surface)


ggplot() +
  geom_sf(data = africa_map, fill = "grey90", color = "grey40")+
  geom_raster(data = continuous_surface,  aes(x = lon, y = lat, fill = value),
              interpolate = TRUE) +
  coord_sf(xlim = c(10, 42), ylim = c(-35, -10), expand = FALSE)
  
  

```

```{r}
ggplot()+
geom_sf(
  data = africa_map,
  fill = "grey90",
  color = "grey40"
)
  
  
  coord_sf(xlim = c(10, 42), ylim = c(-35, -10), expand = FALSE)
```
